---
title: "Global $CO_{2}$ Emissions 1995 and Pressent"
short: "What Keeling missed all these years"
journal: "AER" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
vol: 0
issue: 0
keywords:
  - Replication
  - Modern Science
author:
  - name: Yuna Kim, Jinsoo Chung, Priya Reddy, Jocelyn Thai
 
  
acknowledgements: | 
  The authors would like to thank their instructors from MIDS 271.
  
  
abstract: | 
  Global average temperatures have increased by more than 1℃ since pre-industrial time and the increase of carbon emission has played a key role in the global warming. Human emissions of carbon dioxide and other greenhouse gases are a primary driver of climate change.The rapid warming itself can have significant impacts on climate and natural systems across the world and human's greed for industrialization may have caused it. 
  
header-includes: 

  - '\usepackage{graphicx}'
  - '\usepackage{booktabs}'
output: rticles::aea_article
---

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(tsibble)
library(latex2exp)
library(lubridate)
library(patchwork)
library(magrittr)
library(tsibble)
library(feasts)
library(forecast)

library(sandwich)
library(lmtest)
library(fable)

library(blsR)

library(httr)
library(jsonlite)

library(plyr)
library(dplyr)
library(tidyr)
library(fable)

theme_set(theme_minimal())
knitr::opts_chunk$set(dpi=1000)
```


```{r setup, echo=FALSE}
## default to not show code, unless we ask for it.
knitr::opts_chunk$set(echo=FALSE)
options(digits = 3)
```


Understanding a changing climate, and what it means for the earth's inhabitants is of growing interest to the scientific and policy community. Although, at this point in 1997 it is not entirely clear what the consequences of this growing awareness will be, in this report we present likely outcomes under "business-as-usual" scenarios. In doing so, our hope, is to establish a series of possible futures, and, as evidence, technology, and policy develop over the coming decades, that we can weigh the impacts that carbon-emission reduction efforts might take. 

# Background 
## Carbon Emissions 
We want to understand if atmospheric CO2 levels increased since 1997 and if we anticipate that this trend will continue. High CO2 levels have been shown to negatively play into climate change. The increase in atmospheric CO2 has cascading effects from global warming to ocean acidification. Due to this, we may notice food supply chain disruptions, natural disasters of increasing intensity, and habitat disruption of all ecosystems and the animal species living there. In an economical sense, understanding these patterns is important so we can prepare for any shifts and uncertainities in supply chains that might arise for these new climate patterns. In total understanding the patterns in atmospheric CO2 concentrations is important both so that we can mitigate the existing effects of this increase and prevent continuing upwards trends in CO2.
In addition if our findings are significant, this could persuade other governmental organizations to increase their CO2 monitoring efforts.

## Historical Trends in Atmospheric Carbon 

In 1958 Keeling began continuous monitoring of atmospheric carbon dioxide concentrations from the Mauna Loa Observatory in Hawaii and soon observed a trend increase carbon dioxide levels in addition to the seasonal cycle. He was able to attribute this trend increase to growth in global rates of fossil fuel combustion. This trend has continued to the present, and is known as the "Keeling Curve."

```{r plot the keeling curve, echo = FALSE}
tsibble::as_tsibble(co2) %>%
  ggplot() + 
  aes(x=index, y=value) + 
  geom_line(color = 'steelblue') +
  labs(
    title = TeX(r'(Monthly Mean $CO_2$)'),
    subtitle = 'The "Keeling Curve"',
    x = 'Month and Year',
    y = TeX(r'($CO_2$ parts per million)')
  )
```



# Models and Forecasts 
In this section, we present evaluate two classes models the linear model and the ARIMA model to assess which time series model is most appropriate to use. 

## Data EDA 
 This data measures the mean atmospheric CO2 concentration in Mauna Loa Observatory in Hawaii. The data ranges from 313 parts per million by volume (ppmv) in March 1958 to 406ppmv in November 2018. The data was normalized to remove any influence from local contamination. Carbon dioxide measurements at the Mauna Loa Observatory in Hawaii are made with a type of infrared spectrophotometer, now known as a nondispersive infrared sensor, that is calibrated using World Meteorological Organization standards.


```{r,  echo = FALSE}
co2_ts <- tsibble::as_tsibble(co2) 
colnames(co2_ts)[2] = "co2"
co2_ts['year'] = year(co2_ts$index)
co2_ts['month'] = month(co2_ts$index)
co2_ts
```

```{r EDA ,  echo = FALSE}
p1 <- co2_ts %>% 
  ggplot(aes(x = time(co2), y = co2)) + 
  geom_line() + 
  labs(title = 'CO2 over Time')
p2 <- ggAcf(co2_ts$co2, type = 'correlation') + labs(title = 'ACF of CO2 Levels')
p3 <- ggAcf(co2_ts$co2, type = 'partial')+ labs(title = 'PACF of CO2 Levels')
p4 <- gg_lag(co2, data = co2_ts)
p5 <- co2_ts%>%
  ggplot(aes(x=co2, y= ..count..))+
  geom_histogram()

   (p1 | p2) /
   (p3 | p5)/
    (p4)+
   plot_annotation(
     title    = 'Title',
     subtitle = 'trend by month') & 
   labs(x = NULL, y = 'CO2') 
```

Observations:
- Time Series: From the CO2 Over Time plot, there is an obvious trend of increasing levels throughout time, and we can some regular oscillations which could mean there is seasonality in the data.
- ACF: There's a gradual decay in the ACF values over the lags, perhaps indicating there is a trend in the data. 
- PACF: PACF cuts off to zero after lag 2 and stays that way. However, at lags 12 and 13, the values become significant again. This suggests that although the model exhibits AR model-like behavior, there's some deviances in the data.
- Distribution: 


```{r, echo = FALSE}
co2_yr = co2_ts %>%
  index_by(year) %>%
  summarize(yr_co2 = mean(co2))
#co2_yr
```


```{r, echo = FALSE}
p1 <- co2_yr %>% 
  ggplot(aes(x = year, y = yr_co2)) + 
  geom_line() + 
  labs(title = 'CO2 over Time')
p2 <- ggAcf(co2_yr$yr_co2, type = 'correlation') + labs(title = 'ACF of CO2 Levels')
p3 <- ggAcf(co2_yr$yr_co2, type = 'partial')+ labs(title = 'PACF of CO2 Levels')
p4 <- gg_lag(yr_co2, data = co2_yr)
p5 <- co2_yr%>%
  ggplot(aes(x=yr_co2, y= ..count..))+
  geom_histogram()

   (p1 | p2) /
   (p3 | p5)/
    (p4)+
   plot_annotation(
     title    = 'Title',
     subtitle = 'trend by year') & 
   labs(x = NULL, y = 'CO2') 
```
>Observations:
- Time Series: From the yearly CO2 Over Time plot, there is an obvious trend of increasing levels throughout time, but we can see that there is no seasonality now. This points to the period of the seasonal trend being a year.
- ACF: There's a gradual decay in the ACF values over the lags, indicating there is a still a trend in the data. 



## Linear Models 

To fit linear time trend model to the `co2` series we will compare this to a quadratic time trend model. Discuss whether a logarithmic transformation of the data would be appropriate. Fit a polynomial time trend model that incorporates seasonal dummy variables, and use this model to generate forecasts to the year 2020. 

To begin, we fit a model of the form: 

\begin{equation}
\label{eq:one}
\text{CO}_{2} = \phi_{0} + \phi_{1} + \epsilon_{eit}
\end{equation} 

```{r, echo= FALSE}
#data transform my season
co2_ts <- co2_ts %>%
  mutate(winter = case_when(co2_ts$month %in% c(12,1,2) ~ 1,
                           !co2_ts$month %in% c(12,1,2) ~ 0),
         spring = case_when(co2_ts$month %in% c(3,4,5) ~ 1,
                           !co2_ts$month %in% c(3,4,5) ~ 0),
         summer = case_when(co2_ts$month %in% c(6,7,8) ~ 1,
                           !co2_ts$month %in% c(6,7,8) ~ 0),
         fall = case_when(co2_ts$month %in% c(9,10,11) ~ 1,
                           !co2_ts$month %in% c(9,10,11) ~ 0)
  )
#co2_ts
```

```{r, echo= FALSE}
lin_model <- lm(co2~time(co2), data = co2_ts )
quad_model <-  lm(co2 ~ time(co2) + I(time(co2)**2), data = co2_ts)
poly_model <- lm(co2~time(co2)+I(time(co2)**2)+I(time(co2)**3)+spring+winter+summer+fall, data = co2_ts)

stargazer(lin_model, quad_model, poly_model,)

```


```{r}
#plot model residuals 
mod1_resi <- ggplot(aes(x=lin_model$fitted.values, y=lin_model$residuals), data = lin_model)+
  geom_point()
mod2_resi <- ggplot(aes(x=quad_model$fitted.values, y=quad_model$residuals), data = quad_model)+
  geom_point()
mod3_resi <- ggplot(aes(x=poly_model$fitted.values, y=poly_model$residuals), data = poly_model)+
  geom_point()
```


## Model Forcast

```{r predict 2020}
pred_df <- data.frame("month"=c(1,2,3,4,5,6,7,8,9,10,11,12)/12,
                      "year"=rep(c(2020), each=12),
                      "spring"=c(0,0,1,1,1,0,0,0,0,0,0,0),
                      "winter"=c(1,1,0,0,0,0,0,0,0,0,0,1),
                      "summer"=c(0,0,0,0,0,1,1,1,0,0,0,0),
                      "fall"=c(0,0,0,0,0,0,0,0,1,1,1,0))
pred_df <- pred_df%>%
  mutate(
    "monthyr" = month+year,
    "monthyrsq" = (month+year)**2,
    "monthyrcu" = (month+year)**3,
  )
  
predict(poly_model, newdata= pred_df[c("monthyr","monthyrsq","monthyrcu", 'spring','winter', 'summer', 'fall')])
```

## ARIMA Models 

Sure we also fit some ARIMA models. And talk about them. 

## Forecasts 

```{r}
prediction_1 <- predict(
  object = model_1, 
  newdata = data.frame(x=11), 
  interval = 'prediction'
)
```


Because we have fitted a model, we can make predictions from that model. Our preferred model, named in \autoref{eq:one} is quite simple, and as you might notice, does not in fact match up with the model that we have fitted. However, from this model is is possible to reason about what the outcomes would be if the *input concept* were to be slightly ouside of the observed data range. In particular, if *input concept* were as high as $11$, then we would expect the *output concept* to be `r prediction_1[1,1]`, with a prediction interval that ranges from [`r prediction_1[1,2]`, `r prediction_1[1,3]`]

# Conclusions 

What to conclude is unclear. 

\bibliographystyle{aea}
\bibliography{references}

\appendix
\section{Appendix: Model Robustness}

While the most plausible model that we estimate is reported in the main, "Modeling" section, in this appendix to the article we examine alternative models. Here, our intent is to provide a skeptic that does not accept our assessment of this model as an ARIMA of order (1,2,3) an understanding of model forecasts under alternative scenarios. 

